Ornaments as indicators of Social change before and after European contact at Kiwulan, Northeastern Taiwan

### 2019-08-14 analysis/paper/paper.Rmd
minor changes based on 2nd edit from Ben

```{r, setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/"
)
# prepare for JAS:R
# https://www.elsevier.com/journals/journal-of-archaeological-science-reports/2352-409x/guide-for-authors#2000
# not exceed 6000 words excluding figures, tables, highlights 
# iamge Single column	w = 90 mm 
# image Double column w = 190 mm 
```

@@

Taiwan became part of global trade systems during the early 17th century due to western European expansions that connected Europe and Asia. Despite the colonization of Taiwan by the Spanish and Dutch from 1624 to 1662, there has been little discussion of the archaeology of indigenous responses to the European colonization. @Berrocal2018 examined both archaeological and historical records of Heping Dao in northern Taiwan. This was the location of the colony of San Salvador founded by the Spanish in 1626 and taken over by the Dutch in 1642. Berrocal et al. uncovered parts of the foundation of a Spanish church and a European cemetery that indicate the evidence of European occupation. These remains indicate a direct impact on the local indigenous populations through their modification of the local landscape. In contrast to Heping Dao, Yilan in Northeastern Taiwan is physically isolated and there are no traces of European settlement. This makes Yilan an especially unique location for exploring indirect colonial effects on local indigenous settlements because of isolation. The rugged Hsuehshan Range to the west and Central Range to the south of the Yilan Plain that made it not easily accessible to the European colonists and the colonial control was weaker compared to other parts of this region [@Andrade2007; @Kang2012]. Yilan, also called "Kavalan" by local indigenous people, is an alluvial plain circumscribed by the Pacific and mountains on the sides. Rivers and seas provide the way for the interaction between local indigenous settlements and further afield communities, such as the Basai in northern Taiwan and Han Chinese. Kavalan people offered rice, deer hides, and gold in exchange for beads, metal tools or ornaments, ironwares, porcelains, and textiles with outside traders [@Chen2005; @Hsieh2009; @LiandWu2006].

\Although mountains isolated Yilan by land, the sea provided a way through which indigenous people in Yilan could interact with groups elsewhere, such as indigenous groups, distant traders such as the Han people of China, and Europeans. Large-scale and frequent trading activities brought a wide variety of objects into indigenous communities in Yilan. \The indigenous people of Yilan established their own exchange network that has been described as part of an "inter-insular trade" system [@Chen2005, p. 12] to refer to small-scale regional trade between China and Taiwan on an irregular basis. One of the most commonly traded types of object in this region were personal ornaments such as glass and stone beads. These are one of the most abundant types of foreign goods found at the settlement sites in this region [@Chen2007; @Li2014; @NMTH2005]. Personal adornments in the archaeological record are useful as social signal of an individual's status [@Joyce2005; @Scaramelli2005]. For example, the consumption of stone beads in Southeast Asia during Iron Age is often associated with increasing social stratification or socio-political complexity [@Bellina2014; @Carter2016; @Francis2002; @Theunissen2000; @Kenoyer2000].

>>>>

```{r, setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/"
)
```

@@

Taiwan became part of global trade systems during the early 17th century due to western European expansions that connected Europe and Asia. Despite the colonization of Taiwan by the Spanish and Dutch from 1624 to 1662, there has been little discussion of the archaeology of indigenous responses to the European colonization. @Berrocal2018 examined both archaeological and historical records of Heping Dao in northern Taiwan. This was the location of the colony of San Salvador founded by the Spanish in 1626 and taken over by the Dutch in 1642. Berrocal et al. uncovered parts of the foundation of a Spanish church and a European cemetery that indicate the evidence of European occupation. These remains indicate a direct impact on the local indigenous populations through their modification of the local landscape. In contrast to Heping Dao, Yilan in Northeastern Taiwan is physically isolated and there are no traces of European settlement. This makes Yilan an especially unique location for exploring indirect colonial effects on local indigenous settlements because of isolation. The rugged Hsuehshan Range to the west and Central Range to the south of the Yilan Plain that made it not easily accessible to the European colonists and the colonial control was weaker compared to other parts of this region [@Andrade2007; @Kang2012]. Yilan, also called "Kavalan" by local indigenous people, is an alluvial plain circumscribed by the Pacific and mountains on the sides.

Although mountains isolated Yilan by land, rivers and seas provided a way for the interaction between local indigenous settlements and further afield communities, such as the Basai in northern Taiwan, and Han Chinese. The indigenous people of Yilan established their own exchange network that has been described as part of an "inter-insular trade" system [@Chen2005, p. 12] to refer to small-scale regional trade between China and Taiwan on an irregular basis since 14th century. Kavalan people offered rice, deer hides, and gold in exchange for beads, metal tools or ornaments, ironwares, porcelains, and textiles with outside traders [@Chen2005; @Hsieh2009; @LiandWu2006]. Trading activities brought a wide variety of objects into indigenous communities in Yilan. One of the most commonly traded types of object in this region were personal ornaments such as glass and stone beads. These are one of the most abundant types of foreign goods found at the settlement sites in this region [@Chen2007; @Li2014; @NMTH2005]. Personal adornments in the archaeological record are useful as social signal of an individual's status [@Joyce2005; @Scaramelli2005]. For example, the consumption of stone beads in Southeast Asia during Iron Age is often associated with increasing social stratification or socio-political complexity [@Bellina2014; @Carter2016; @Francis2002; @Theunissen2000; @Kenoyer2000].
